---
title: "Honey Jags 2"
author: "Calvin Fertig III"
date: "2026-02-9"
output: pdf_document
---



```{r, warning = FALSE}
library(rjags)
library(jagsUI)
library(ggmcmc)
df <- read.csv("honey_bee_state.csv")
colnames(df)

df$Year  <- as.numeric(scale(df$Year))
df$VARROAMITES <- as.numeric(scale(df$VARROAMITES))
df$PESTICIDES  <- as.numeric(scale(df$PESTICIDES))
df$TemperatureC <- as.numeric(scale(df$TemperatureC))
df$Percipitation.Inches <- as.numeric(scale(df$Percipitation.Inches))
df$DISEASE <- as.numeric(scale(df$DISEASE))
source("sampleHDI.R")
df$state_id <- as.numeric(factor(df$State))
```




# Slope and Intercept Heir, all other not



```{r}
mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0[state[n]] + beta1[state[n]] * Year[n] + beta2*Mites[n] + beta3*Pesti[n] + beta4*Temp[n] + beta5*Perc[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Other Data Points
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)
  beta5 ~ dnorm(0, 0.001)
  
  
  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)


  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c(
  "beta0",      # state intercepts
  "beta1",      # state slopes
  "mu_beta1",   # mean slope
  "sigma_beta1",
  "sigma_y",
  "beta2",
  "beta3",
  "beta4",
  "beta5"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,      
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_STSL_Only_No_Inter <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```





```{r}
Full_Fitted_Model_Hier_STSL_Only_No_Inter

traceplot(Full_Fitted_Model_Hier_STSL_Only_No_Inter)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_STSL_Only_No_Inter$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```

## No intercept heir, slope hier, no other

```{r}
mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0 + beta1[state[n]] * Year[n] + beta2*Mites[n] + beta3*Pesti[n] + beta4*Temp[n] + beta5*Perc[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Other Data Points
  beta0 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)
  beta5 ~ dnorm(0, 0.001)
  
  
  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)


  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c(
  "beta0",      # intercept
  "beta1",      # state slopes
  "mu_beta1",   # mean slope
  "sigma_beta1",
  "sigma_y",
  "beta2",
  "beta3",
  "beta4",
  "beta5"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,      
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_SL_Only_No_Inter <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```


```{r}
Full_Fitted_Model_Hier_SL_Only_No_Inter
```



```{r}
traceplot(Full_Fitted_Model_Hier_SL_Only_No_Inter)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_SL_Only_No_Inter$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```






# slope and intercept heir, none other, but with interaction 


```{r}
mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0[state[n]] + beta1[state[n]] * Year[n] + beta2*Mites[n] + beta3*Pesti[n] + beta4*Temp[n] + beta5*Perc[n] +             
         beta6*Mites[n]*Pesti[n] +
         beta7*Mites[n]*Temp[n] +
         beta8*Mites[n]*Perc[n] +
         beta9*Pesti[n]*Temp[n] +
         beta10*Pesti[n]*Perc[n] +
         beta11*Temp[n]*Perc[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Other Data Points
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)
  beta5 ~ dnorm(0, 0.001)
  beta6  ~ dnorm(0, 0.001)
  beta7  ~ dnorm(0, 0.001)
  beta8  ~ dnorm(0, 0.001)
  beta9  ~ dnorm(0, 0.001)
  beta10 ~ dnorm(0, 0.001)
  beta11 ~ dnorm(0, 0.001)
  
  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)


  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c(
  "beta0",      # state intercepts
  "beta1",      # state slopes
  "mu_beta1",   # mean slope
  "sigma_beta1",
  "sigma_y",
  "beta2",
  "beta3",
  "beta4",
  "beta5",
  "beta6",
  "beta7",
  "beta8",
  "beta9",
  "beta10",
  "beta11"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,      
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_STSL_Only_with_Inter <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```

```{r}
Full_Fitted_Model_Hier_STSL_Only_with_Inter
```






```{r}
traceplot(Full_Fitted_Model_Hier_STSL_Only_with_Inter)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_STSL_Only_with_Inter$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```



# Full Model, Hier on slope and intercept, ALL interactions

```{r}

mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {

    Y[n] ~ dnorm(mu[n], tau_y)

    mu[n] <- beta0[state[n]] +
             beta1[state[n]] * Year[n] +

             # Main effects
             beta[2]  * Mites[n] +
             beta[3]  * Pesti[n] +
             beta[4]  * Temp[n]  +
             beta[5]  * Perc[n]  +

             # Two-way
             beta[6]  * Mites[n] * Pesti[n] +
             beta[7]  * Mites[n] * Temp[n]  +
             beta[8]  * Mites[n] * Perc[n]  +
             beta[9]  * Pesti[n] * Temp[n]  +
             beta[10] * Pesti[n] * Perc[n]  +
             beta[11] * Temp[n]  * Perc[n]  +

             # Three-way
             beta[12] * Mites[n] * Pesti[n] * Temp[n] +
             beta[13] * Mites[n] * Pesti[n] * Perc[n] +
             beta[14] * Mites[n] * Temp[n]  * Perc[n] +
             beta[15] * Pesti[n] * Temp[n]  * Perc[n] +

             # Four-way
             beta[16] * Mites[n] * Pesti[n] * Temp[n] * Perc[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Priors 
  for (k in 2:16) {
    beta[k] ~ dnorm(0, 0.001)
  }

  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)

  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"

params <- c(
  "beta0",
  "beta1",
  "mu_beta0",
  "sigma_beta0",
  "mu_beta1",
  "sigma_beta1",
  "sigma_y",
  "beta"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_STSL_All_Interactions <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3,
  n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000,
  n.thin = 10,
  parallel = FALSE,
  DIC = TRUE
)

```

```{r}
Full_Fitted_Model_Hier_STSL_All_Interactions
```




```{r}
traceplot(Full_Fitted_Model_Hier_STSL_All_Interactions)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_STSL_All_Interactions$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```



# Heiracrhy only slope, but with Full interactions


```{r}
mod_hier_slope_only <- "
model {

  # Likelihood
  for (n in 1:N) {

    Y[n] ~ dnorm(mu[n], tau_y)

    mu[n] <- beta0 +
             beta1[state[n]] * Year[n] +

             # Main effects
             beta[2]  * Mites[n] +
             beta[3]  * Pesti[n] +
             beta[4]  * Temp[n]  +
             beta[5]  * Perc[n]  +

             # Two-way interactions
             beta[6]  * Mites[n] * Pesti[n] +
             beta[7]  * Mites[n] * Temp[n]  +
             beta[8]  * Mites[n] * Perc[n]  +
             beta[9]  * Pesti[n] * Temp[n]  +
             beta[10] * Pesti[n] * Perc[n]  +
             beta[11] * Temp[n]  * Perc[n]  +

             # Three-way interactions
             beta[12] * Mites[n] * Pesti[n] * Temp[n] +
             beta[13] * Mites[n] * Pesti[n] * Perc[n] +
             beta[14] * Mites[n] * Temp[n]  * Perc[n] +
             beta[15] * Pesti[n] * Temp[n]  * Perc[n] +

             # Four-way interaction
             beta[16] * Mites[n] * Pesti[n] * Temp[n] * Perc[n]
  }

  # Global intercept
  beta0 ~ dnorm(0, 0.001)

  # Hierarchical slope for Year
  for (s in 1:S) {
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Priors 
  for (k in 2:16) {
    beta[k] ~ dnorm(0, 0.001)
  }

  # Hyperpriors for slope hierarchy
  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)

  # Residual variance
  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"

params <- c(
  "beta0",
  "beta1",
  "mu_beta1",
  "sigma_beta1",
  "beta",
  "sigma_y"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_Slope_Only <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_slope_only),
  n.chains = 3,
  n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000,
  n.thin = 10,
  parallel = FALSE,
  DIC = TRUE
)

```


```{r}
Full_Fitted_Model_Hier_Slope_Only
```







```{r}
traceplot(Full_Fitted_Model_Hier_Slope_Only)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_Slope_Only$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```






# Slope and inter hierachy, no interactions but a few disease added.



```{r}
mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0[state[n]] + beta1[state[n]] * Year[n] + beta2*Mites[n] + beta3*Pesti[n] + beta4*Temp[n] + beta5*Perc[n] + beta6*Di[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Other Data Points
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)
  beta4 ~ dnorm(0, 0.001)
  beta5 ~ dnorm(0, 0.001)
  beta6 ~ dnorm(0, 0.001) 
  
  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)


  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c(
  "beta0",      # state intercepts
  "beta1",      # state slopes
  "mu_beta1",   # mean slope
  "sigma_beta1",
  "sigma_y",
  "beta2",
  "beta3",
  "beta4",
  "beta5",
  "beta6"
)

inits <- NULL

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,      
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches,
  Di    = df$DISEASE
)

Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```


```{r}
Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE
```







```{r}
traceplot(Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE)
par(mfrow=c(3,3))
autocorr.plot(Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```







# Heirarchy on all variables, no interaction.





```{r}
mod_hier_all <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)

    mu[n] <- beta0[state[n]] +
             beta1[state[n]] * Year[n] +
             beta2[state[n]] * Mites[n] +
             beta3[state[n]] * Pesti[n] +
             beta4[state[n]] * Temp[n] +
             beta5[state[n]] * Perc[n]
  }

  # State-level coefficients
  for (s in 1:S) {

    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
    beta2[s] ~ dnorm(mu_beta2, tau_beta2)
    beta3[s] ~ dnorm(mu_beta3, tau_beta3)
    beta4[s] ~ dnorm(mu_beta4, tau_beta4)
    beta5[s] ~ dnorm(mu_beta5, tau_beta5)
  }

  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)

  mu_beta2 ~ dnorm(0, 0.001)
  sigma_beta2 ~ dunif(0, 50)
  tau_beta2 <- pow(sigma_beta2, -2)

  mu_beta3 ~ dnorm(0, 0.001)
  sigma_beta3 ~ dunif(0, 50)
  tau_beta3 <- pow(sigma_beta3, -2)

  mu_beta4 ~ dnorm(0, 0.001)
  sigma_beta4 ~ dunif(0, 50)
  tau_beta4 <- pow(sigma_beta4, -2)

  mu_beta5 ~ dnorm(0, 0.001)
  sigma_beta5 ~ dunif(0, 50)
  tau_beta5 <- pow(sigma_beta5, -2)

  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)
}
"


params <- c(
  "beta0","beta1","beta2","beta3","beta4","beta5",
  "mu_beta0","mu_beta1","mu_beta2","mu_beta3","mu_beta4","mu_beta5",
  "sigma_beta0","sigma_beta1","sigma_beta2","sigma_beta3","sigma_beta4","sigma_beta5",
  "sigma_y"
)


jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id)),
  Mites = df$VARROAMITES,
  Pesti = df$PESTICIDES,
  Temp  = df$TemperatureC,
  Perc  = df$Percipitation.Inches
)

Full_Fitted_Model_Hier_All <- jags(
  data = jags_data,
  inits = NULL,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_all),
  n.chains = 3,
  n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000,
  n.thin = 10,
  parallel = FALSE,
  DIC = TRUE
)
```



```{r}
Full_Fitted_Model_Hier_All
```



```{r}
traceplot(Full_Fitted_Model_Hier_All)
par(mfrow=c(3,3))
#autocorr.plot(Full_Fitted_Model_Hier_All$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```


# No other predictors

```{r}
mod_hier_both <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0[state[n]] + beta1[state[n]] * Year[n]
  }

  # State intercepts and slopes
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # Hyperpriors
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)


  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c(
  "beta0",      # state intercepts
  "beta1",      # state slopes
  "mu_beta1",   # mean slope
  "sigma_beta1",
  "sigma_y"
)


jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,      
  state = df$state_id,
  N     = nrow(df),
  S     = length(unique(df$state_id))
)

Slope_Inter_Hier_Only <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_both),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```



```{r}
Slope_Inter_Hier_Only
```







```{r}
traceplot(Slope_Inter_Hier_Only)
par(mfrow=c(3,3))
autocorr.plot(Slope_Inter_Hier_Only$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```


# only INtercept


```{r}
mod_hier_intercept <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0[state[n]] + beta1 * Year[n]
  }

  # State intercepts
  for (s in 1:S) {
    beta0[s] ~ dnorm(mu_beta0, tau_beta0)
  }

  # Hyperpriors for intercepts
  mu_beta0 ~ dnorm(0, 0.001)
  sigma_beta0 ~ dunif(0, 50)
  tau_beta0 <- pow(sigma_beta0, -2)

  #  slope
  beta1 ~ dnorm(0, 0.001)

  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"


params <- c(
  "beta0",      # stateintercepts
  "beta1",      #  slope
  "sigma_y"
)

jags_data <- list(
  Y     = df$HpH,
  Year  = df$Year,        
  state = df$state_id,    
  N     = nrow(df),
  S     = length(unique(df$state_id))
)



Only_Inter <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier_intercept),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)

```




```{r}
Only_Inter
```







```{r}
traceplot(Only_Inter)
par(mfrow=c(3,3))
autocorr.plot(Only_Inter$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```
















# Only Slope Hier


```{r}
mod_hier <- "
model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau_y)
    mu[n] <- beta0 + beta1[state[n]] * Year[n]
  }

  # State slopes
  for (s in 1:S) {
    beta1[s] ~ dnorm(mu_beta1, tau_beta1)
  }

  # priors
  mu_beta1 ~ dnorm(0, 0.001)
  sigma_beta1 ~ dunif(0, 50)
  tau_beta1 <- pow(sigma_beta1, -2)

  # intercept
  beta0 ~ dnorm(0, 0.001)

  sigma_y ~ dunif(0, 100)
  tau_y <- pow(sigma_y, -2)

}
"
params <- c('beta0', 'beta1', 'mu_beta1', 'sigma_beta1')

SLOPE_HIER_Only <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(mod_hier),
  n.chains = 3, n.adapt = 5000,
  n.iter = 101000,
  n.burnin = 1000, n.thin = 10,
  parallel = FALSE, DIC = TRUE
)
```











```{r}
SLOPE_HIER_Only
```







```{r}
traceplot(SLOPE_HIER_Only)
par(mfrow=c(3,3))
autocorr.plot(SLOPE_HIER_Only$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```








# No Hier

```{r}
mod <- "

model {

  # Likelihood
  for (n in 1:N) {
    Y[n] ~ dnorm(mu[n], tau)
    mu[n] <- beta0 + beta1 * Year[n]
  }

  # Priors
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)

  sigma ~ dunif(0, 100)
  tau <- pow(sigma, -2)

}"


jags_data <- list(
  Y    = df$HpH,
  Year = df$Year,   # must be 1–8
  N    = nrow(df)
)
inits <- NULL
params <- c('beta0', 'beta1')


## Run the Bayesian model using MCMC
No_Hier <- jags(data = jags_data, inits = inits,
               parameters.to.save = params,
               model.file = textConnection(mod),
               n.chains = 3, n.adapt = 5000,
               n.iter = 101000,
               n.burnin = 1000, n.thin = 10,
               parallel = FALSE, DIC = TRUE)
```





```{r}
No_Hier
```







```{r}
traceplot(No_Hier)
par(mfrow=c(3,3))
autocorr.plot(No_Hier$samples,lag.max=40,auto.layout=FALSE,ask=FALSE)
```












```{r}
# Put all model objects into a named list
model_list <- list(
  "Full_Fitted_Model_Hier_STSL_Only_No_Inter" = Full_Fitted_Model_Hier_STSL_Only_No_Inter,
  "Full_Fitted_Model_Hier_SL_Only_No_Inter" = Full_Fitted_Model_Hier_SL_Only_No_Inter,
  "Full_Fitted_Model_Hier_STSL_Only_with_Inter" = Full_Fitted_Model_Hier_STSL_Only_with_Inter,
  "Full_Fitted_Model_Hier_STSL_All_Interactions" = Full_Fitted_Model_Hier_STSL_All_Interactions,
  "Full_Fitted_Model_Hier_Slope_Only" = Full_Fitted_Model_Hier_Slope_Only,
  "Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE" = Full_Fitted_Model_Hier_STSL_Only_No_Inter_and_DISEASE,
  "Full_Fitted_Model_Hier_All" = Full_Fitted_Model_Hier_All,
  "Slope_Inter_Hier_Only" = Slope_Inter_Hier_Only,
  "Only_Inter" = Only_Inter,
  "SLOPE_HIER_Only" = SLOPE_HIER_Only
)

# Extract DIC values safely (handles models without DIC)
dic_values <- sapply(model_list, function(mod) {
  if (!is.null(mod$DIC)) {
    return(mod$DIC)
  } else {
    return(NA)
  }
})

# Create comparison table
dic_table <- data.frame(
  Model = names(dic_values),
  DIC = as.numeric(dic_values)
)

# Sort by DIC (lowest is best)
dic_table <- dic_table[order(dic_table$DIC), ]

# Print nicely
print(dic_table)

# Optional: nicer display in R Markdown
knitr::kable(dic_table, digits = 3)

```




```{r}
lme4::lmer(HpH~ (1|State) + Year, data = df)
```






